---
title: "Tutorial 4: Comparing models (estimating K)"
author: "Bob Verity"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{"Comparing models (estimating K)"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
set.seed(3)
library(MALECOT)
reps <- 1e2
```

This vignette demonstrates model comparison in MALECOT. It covers:

1. Running thermodynamic MCMC
2. Checking convergence of thermodynamic MCMC
3. Estimating *K* via generalized thermodynamic integration (GTI)
4. Comparing parameter sets through the model evidence

This tutorial assumes some prior knowledge about MALECOT, so if you are completely new to the program we recommend working through the simpler [bi-allelic tutorial](https://bobverity.github.io/MALECOT/articles/tutorial-biallelic.html) first.

## Background on Bayesian model comparison

When carrying out Bayesian analysis it is useful to make a distinction between two types of analysis:

1. parameter estimation within a model
2. comparison between models

As an example, we might create a model $\mathcal{M}$ in which we write down the probability of the observed data $x$ as a function of the unknown allele frequencies $p$. In other words, we write down the likelihood $\mbox{Pr}(x \: | \: p, \mathcal{M})$.

In Bayesian parameter estimation we are trying to get at the posterior probability $\mbox{Pr}(p \: | \: x, \mathcal{M})$. We typically do this using MCMC, which produces a series of draws from the posterior distribution.

But what if we want to know the posterior probability of the *model*, rather than the parameters of the model? In other words, we want to know $\mbox{Pr}(\mathcal{M} \: | \: x)$. Calculating this quantity requires that we integrate over all unknown parameters:

$$\mbox{Pr}(\mathcal{M} \: | \: x) = \int \mbox{Pr}(\mathcal{M}, p \: | \: x) dp$$

This is often an extremely high-dimensional integral - for example in MALECOT there could easily be hundreds of unknown parameters - making it computationally infeasible by most methods. Regular MCMC cannot help us here either, because it only produces draws from the posterior distribution rather than normalised values.

One way around this is to use an advanced MCMC technique known as thermodynamic integration (TI). In TI we run multiple MCMC chains, each at a different "rung" on a temperature ladder. The hotter the chain, the flatter the target distribution. The log-likelihoods over all chains are then combined in a single calculation that - by what can only be described as mathematical magic! - is asymptotically equal to $\log[\mbox{Pr}(x \: | \: \mathcal{M})]$. We can then apply a prior over models, for example giving each model equal weight, to arrive at the desired posterior value $\mbox{Pr}(\mathcal{M} \: | \: x)$. *Generalised* thermodynamic integration (GTI) differs from regular TI in that it uses a slightly different calculation when combining information across rungs that leads to lower bias and higher precision. 

Hopefully this is enough background to run thermodynamic MCMC in MALECOT and understand the results. For those eager to understand all the mathematical details, see [this vignette](https://bobverity.github.io/MALECOT/articles/gti-mathematical-details.html).

## Running a thermodynamic MCMC

For the sake of this tutorial we will use simulated bi-allelic data drawn from $K = 3$ subpopulations. We create a new project and bind this data:

```{r}
# simulate data
mysim <- sim_data(data_format = "biallelic", n = 100, L = 24, K = 3)
```

```{r, echo=FALSE}
# secretly save/load data from file so results are consistent
#saveRDS(mysim, file = "../inst/extdata/tutorial4_mysim.rds")
mysim <- malecot_file("tutorial4_mysim.rds")
```

```{r}
# create project and bind data
myproj <- malecot_project()
myproj <- bind_data_biallelic(myproj, df = mysim$data, ID_col = 1, pop_col = 2)
```

For our first parameter set we will assume a very basic model with default parameters; which means a Poisson prior on COI, a flat prior on allele frequencies and no error estimation.

```{r}
# create parameter set
myproj <- new_set(myproj, name = "simple model")
```

We will then run the MCMC for values of $K$ from 1 to 5. To make this thermodymic MCMC we use the `rungs` argument to set the number of rungs on the temperature ladder, here opting for 10 rungs. Be warned - this MCMC will take considerably longer to run than regular MCMC, so be prepared to go make yourself a cup of tea!

```{r}
# run thermodynamic MCMC
myproj <- run_mcmc(myproj, K = 1:5, burnin = reps, converge_test = 1e2,
                   samples = reps, rungs = 10, pb_markdown =  TRUE)
```

Notice that convergence times are much longer than under regular MCMC. This is bacause the MCMC is only deemed to have converged when *every* chain has converged, meaning we are as weak as the weakest link. When running thermodynamic MCMC it is therefore important to keep an eye on convergence, and increase the number of burn-in iterations if needed.

## Checking thermodynamic MCMC behaviour

There are more moving parts in thermodynamic MCMC, which on the one hand can help it to explore the space well, but on the other hand means we have more things to check.

First, we should perform the same diagnostic checks as for regular MCMC, ensuring that the cold chain has explored the space well:

```{r}
plot_loglike_dignostic(myproj, K = 3)
```

Running `get_ESS()` on thermodynamic MCMC output now prints the effective sample size of every MCMC rung:

```{r}
get_ESS(myproj, K = 3)
```

We interpret ESS for hot chains the same way as for the cold chain - this is the number of samples that we have obtained from the target distribution once autocorrelation has been taken into account. Therefore if we see small values (tens to hundreds as a rule of thumb) then we should be concerned that our thermodynamic MCMC has not explored the space well, and we should repeat the analysis with a larger number of samples.

Another useful plotting function is `plot_loglike()`, which plots the 95% quantiles of the log-likelihood under each temperature rung:

```{r}
plot_loglike(myproj, K = 3)
```

This plot should be always-increasing from left to right, aside from small variations due to the random nature of MCMC. If any rungs stand out from the overall pattern then it is worth running `plot_loglike_diagnostic()` on these particular rungs using the `rung` argument, for example:

```{r}
plot_GTI_path(myproj, K = 3)
```

```{r}
plot_logevidence_K(myproj)
```

```{r}
plot_posterior_K(myproj)
```


## Comparing different parameter sets

```{r}
# create parameter set
myproj <- new_set(myproj, name = "error model", estimate_error = TRUE)

myproj
```

```{r}
# run MCMC
myproj <- run_mcmc(myproj, K = 1:5, burnin = reps, converge_test = 1e2,
                   samples = reps, rungs = 10, pb_markdown =  TRUE)
```

```{r}
plot_loglike_dignostic(myproj, K = 3)
```

```{r, fig.height=3, fig.width=4}
plot_e(myproj, K = 3)
```

```{r, fig.height=5, fig.width=8}
plot_loglike(myproj, K = 3)

plot_GTI_path(myproj, K = 3)
```


```{r}
plot_logevidence_K(myproj)
```

```{r}
plot_posterior_K(myproj)
```

```{r}
plot_logevidence_model(myproj)
```

```{r}
plot_posterior_model(myproj)
```

