---
title: "Power analysis - within model"
author: "Bob Verity"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{"Power analysis - within model"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
set.seed(1)
library(MALECOT)
```

One of the major advantages of simulated data is that we can test the power of the program under different scenarios and data sizes, hopefully allowing us to design field and lab studies that are powered to detect a signal if it is there. MALECOT is Bayesian and therefore power in this context is a Bayesian equivalent of traditional power, defined as the posterior probability of the true value averaged over a large number of simulations from the same model. For example, if the power to detect population structure is 0.9 then that means the true grouping will tend to have a posterior probability of 0.9.

This vignette explores power to detect population structure for the following parameter ranges:

* Always assuming $K = 5$ subpopulations
* Sample size (*n*) in the range $[25, 100]$. Note that this is the total sample size over all 5 subpopulations, meaning the number of samples per subpopulation is 1/5th this value
* Loci (*L*) in the range $[10, 100]$
* Mean COI per subpopulation in ${1.2, 2.0, 5.0}$, representing low, moderate and high transmission intensity. The assumed COI distribution is Poisson
* Shape parameter of the prior on allele frequencies (*lambda1*) in $\{1, 5, 10\}$, representing different levels of skewed allele frequency distribution
* Proportion erronious genotyping calls (both false homozygote and false heterozygote) in $\{0.00, 0.05, 0.10\}$

Results are averaged over 50 simulated data sets, analysed by MCMC with the following parameters:

* 10,000 burn-in iterations. Test for convergence automatically every 100 iterations
* 10,000 sampling iterations
* Single temperature rung (no thermodynamic MCMC)

The three priors on allele frequencies correspond to the following distributions:

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=5}
library(tidyr)
library(plotly)

# load power analysis results
df <- malecot_file("power_within_results_mean.rds")
df$power_structure <- round(df$power_structure, 2)

# get loci into wide format
df <- spread(df, L, power_structure)
names(df)[-(1:4)] <- paste0("L", names(df)[-(1:4)])

# define vectors of values
COI_mean <- c(1.2, 2, 5)
loci <- c(10, 20, 50, 100)
```

```{r, echo=FALSE, fig.width=10, fig.height=4}
library(gridExtra)

# produce prior plots
plot1 <- prior_plot <- plot_prior_p(lambda = c(1, 1)) + ggtitle("lambda1 = 1")
plot2 <- prior_plot <- plot_prior_p(lambda = c(5, 1)) + ggtitle("lambda1 = 5")
plot3 <- prior_plot <- plot_prior_p(lambda = c(10, 1)) + ggtitle("lambda1 = 10")

grid.arrange(plot1, plot2, plot3, nrow = 1)
```

## Power to detect population structure

------

### Error = 0

<br>
<br>

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=5}
# choose parameters to plot
error <- 0
lambda1 <- 1

# create basic plot and add multiple traces
p <- plot_ly()
for (i in 1:3) {
  df_sub <- df[df$error == error & df$lambda1 == lambda1 & df$COI_mean == COI_mean[i],]
  for (j in 1:4) {
    p <- add_trace(p, 
                 x = unique(df$n), 
                 y = df_sub[,j+4], 
                 type = 'scatter', 
                 mode = 'lines+markers',
                 visible = i == 1,
                 name = paste0("loci = ", loci[j]))
  }
}

# create dropdown buttons as list
button_list <- list()
for (i in 1:3) {
  v <- rep(FALSE,3)
  v[i] <- TRUE
  button_list[[i]] <- list(method = "restyle",
                       args = list("visible", as.list(rep(v, each = 4))),
                       label = paste0("COI_mean = ", COI_mean[i]))
}

# finalise layout plto plot
p <- layout(p,
            title = "lambda1 = 1",
            yaxis = list(title = "probability correct assignment", range = c(0,1)),
            xaxis = list(title = "sample size", range = c(0,250)),
            updatemenus = list(
              list(
                yanchor = 'auto',
                buttons = button_list
                )
              )
            )

# print plot
p
```

<br>
<br>

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=5}
# choose parameters to plot
error <- 0
lambda1 <- 5

# create basic plot and add multiple traces
p <- plot_ly()
for (i in 1:3) {
  df_sub <- df[df$error == error & df$lambda1 == lambda1 & df$COI_mean == COI_mean[i],]
  for (j in 1:4) {
    p <- add_trace(p, 
                 x = unique(df$n), 
                 y = df_sub[,j+4], 
                 type = 'scatter', 
                 mode = 'lines+markers',
                 visible = i == 1,
                 name = paste0("loci = ", loci[j]))
  }
}

# create dropdown buttons as list
button_list <- list()
for (i in 1:3) {
  v <- rep(FALSE,3)
  v[i] <- TRUE
  button_list[[i]] <- list(method = "restyle",
                       args = list("visible", as.list(rep(v, each = 4))),
                       label = paste0("COI_mean = ", COI_mean[i]))
}

# finalise layout plto plot
p <- layout(p,
            title = "lambda1 = 5",
            yaxis = list(title = "probability correct assignment", range = c(0,1)),
            xaxis = list(title = "sample size", range = c(0,250)),
            updatemenus = list(
              list(
                yanchor = 'auto',
                buttons = button_list
                )
              )
            )

# print plot
p
```

<br>
<br>

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=5}
# choose parameters to plot
error <- 0
lambda1 <- 10

# create basic plot and add multiple traces
p <- plot_ly()
for (i in 1:3) {
  df_sub <- df[df$error == error & df$lambda1 == lambda1 & df$COI_mean == COI_mean[i],]
  for (j in 1:4) {
    p <- add_trace(p, 
                 x = unique(df$n), 
                 y = df_sub[,j+4], 
                 type = 'scatter', 
                 mode = 'lines+markers',
                 visible = i == 1,
                 name = paste0("loci = ", loci[j]))
  }
}

# create dropdown buttons as list
button_list <- list()
for (i in 1:3) {
  v <- rep(FALSE,3)
  v[i] <- TRUE
  button_list[[i]] <- list(method = "restyle",
                       args = list("visible", as.list(rep(v, each = 4))),
                       label = paste0("COI_mean = ", COI_mean[i]))
}

# finalise layout plto plot
p <- layout(p,
            title = "lambda1 = 10",
            yaxis = list(title = "probability correct assignment", range = c(0,1)),
            xaxis = list(title = "sample size", range = c(0,250)),
            updatemenus = list(
              list(
                yanchor = 'auto',
                buttons = button_list
                )
              )
            )

# print plot
p
```

<br>
<br>

------

### Error = 0.05

<br>
<br>

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=5}
# choose parameters to plot
error <- 0.05
lambda1 <- 1

# create basic plot and add multiple traces
p <- plot_ly()
for (i in 1:3) {
  df_sub <- df[df$error == error & df$lambda1 == lambda1 & df$COI_mean == COI_mean[i],]
  for (j in 1:4) {
    p <- add_trace(p, 
                 x = unique(df$n), 
                 y = df_sub[,j+4], 
                 type = 'scatter', 
                 mode = 'lines+markers',
                 visible = i == 1,
                 name = paste0("loci = ", loci[j]))
  }
}

# create dropdown buttons as list
button_list <- list()
for (i in 1:3) {
  v <- rep(FALSE,3)
  v[i] <- TRUE
  button_list[[i]] <- list(method = "restyle",
                       args = list("visible", as.list(rep(v, each = 4))),
                       label = paste0("COI_mean = ", COI_mean[i]))
}

# finalise layout plto plot
p <- layout(p,
            title = "lambda1 = 1",
            yaxis = list(title = "probability correct assignment", range = c(0,1)),
            xaxis = list(title = "sample size", range = c(0,250)),
            updatemenus = list(
              list(
                yanchor = 'auto',
                buttons = button_list
                )
              )
            )

# print plot
p
```

<br>
<br>

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=5}
# choose parameters to plot
error <- 0.05
lambda1 <- 5

# create basic plot and add multiple traces
p <- plot_ly()
for (i in 1:3) {
  df_sub <- df[df$error == error & df$lambda1 == lambda1 & df$COI_mean == COI_mean[i],]
  for (j in 1:4) {
    p <- add_trace(p, 
                 x = unique(df$n), 
                 y = df_sub[,j+4], 
                 type = 'scatter', 
                 mode = 'lines+markers',
                 visible = i == 1,
                 name = paste0("loci = ", loci[j]))
  }
}

# create dropdown buttons as list
button_list <- list()
for (i in 1:3) {
  v <- rep(FALSE,3)
  v[i] <- TRUE
  button_list[[i]] <- list(method = "restyle",
                       args = list("visible", as.list(rep(v, each = 4))),
                       label = paste0("COI_mean = ", COI_mean[i]))
}

# finalise layout plto plot
p <- layout(p,
            title = "lambda1 = 5",
            yaxis = list(title = "probability correct assignment", range = c(0,1)),
            xaxis = list(title = "sample size", range = c(0,250)),
            updatemenus = list(
              list(
                yanchor = 'auto',
                buttons = button_list
                )
              )
            )

# print plot
p
```

<br>
<br>

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=5}
# choose parameters to plot
error <- 0.05
lambda1 <- 10

# create basic plot and add multiple traces
p <- plot_ly()
for (i in 1:3) {
  df_sub <- df[df$error == error & df$lambda1 == lambda1 & df$COI_mean == COI_mean[i],]
  for (j in 1:4) {
    p <- add_trace(p, 
                 x = unique(df$n), 
                 y = df_sub[,j+4], 
                 type = 'scatter', 
                 mode = 'lines+markers',
                 visible = i == 1,
                 name = paste0("loci = ", loci[j]))
  }
}

# create dropdown buttons as list
button_list <- list()
for (i in 1:3) {
  v <- rep(FALSE,3)
  v[i] <- TRUE
  button_list[[i]] <- list(method = "restyle",
                       args = list("visible", as.list(rep(v, each = 4))),
                       label = paste0("COI_mean = ", COI_mean[i]))
}

# finalise layout plto plot
p <- layout(p,
            title = "lambda1 = 10",
            yaxis = list(title = "probability correct assignment", range = c(0,1)),
            xaxis = list(title = "sample size", range = c(0,250)),
            updatemenus = list(
              list(
                yanchor = 'auto',
                buttons = button_list
                )
              )
            )

# print plot
p
```

------

### Error = 0.10

<br>
<br>

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=5}
# choose parameters to plot
error <- 0.10
lambda1 <- 1

# create basic plot and add multiple traces
p <- plot_ly()
for (i in 1:3) {
  df_sub <- df[df$error == error & df$lambda1 == lambda1 & df$COI_mean == COI_mean[i],]
  for (j in 1:4) {
    p <- add_trace(p, 
                 x = unique(df$n), 
                 y = df_sub[,j+4], 
                 type = 'scatter', 
                 mode = 'lines+markers',
                 visible = i == 1,
                 name = paste0("loci = ", loci[j]))
  }
}

# create dropdown buttons as list
button_list <- list()
for (i in 1:3) {
  v <- rep(FALSE,3)
  v[i] <- TRUE
  button_list[[i]] <- list(method = "restyle",
                       args = list("visible", as.list(rep(v, each = 4))),
                       label = paste0("COI_mean = ", COI_mean[i]))
}

# finalise layout plto plot
p <- layout(p,
            title = "lambda1 = 1",
            yaxis = list(title = "probability correct assignment", range = c(0,1)),
            xaxis = list(title = "sample size", range = c(0,250)),
            updatemenus = list(
              list(
                yanchor = 'auto',
                buttons = button_list
                )
              )
            )

# print plot
p
```

<br>
<br>

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=5}
# choose parameters to plot
error <- 0.10
lambda1 <- 5

# create basic plot and add multiple traces
p <- plot_ly()
for (i in 1:3) {
  df_sub <- df[df$error == error & df$lambda1 == lambda1 & df$COI_mean == COI_mean[i],]
  for (j in 1:4) {
    p <- add_trace(p, 
                 x = unique(df$n), 
                 y = df_sub[,j+4], 
                 type = 'scatter', 
                 mode = 'lines+markers',
                 visible = i == 1,
                 name = paste0("loci = ", loci[j]))
  }
}

# create dropdown buttons as list
button_list <- list()
for (i in 1:3) {
  v <- rep(FALSE,3)
  v[i] <- TRUE
  button_list[[i]] <- list(method = "restyle",
                       args = list("visible", as.list(rep(v, each = 4))),
                       label = paste0("COI_mean = ", COI_mean[i]))
}

# finalise layout plto plot
p <- layout(p,
            title = "lambda1 = 5",
            yaxis = list(title = "probability correct assignment", range = c(0,1)),
            xaxis = list(title = "sample size", range = c(0,250)),
            updatemenus = list(
              list(
                yanchor = 'auto',
                buttons = button_list
                )
              )
            )

# print plot
p
```

<br>
<br>

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=5}
# choose parameters to plot
error <- 0.10
lambda1 <- 10

# create basic plot and add multiple traces
p <- plot_ly()
for (i in 1:3) {
  df_sub <- df[df$error == error & df$lambda1 == lambda1 & df$COI_mean == COI_mean[i],]
  for (j in 1:4) {
    p <- add_trace(p, 
                 x = unique(df$n), 
                 y = df_sub[,j+4], 
                 type = 'scatter', 
                 mode = 'lines+markers',
                 visible = i == 1,
                 name = paste0("loci = ", loci[j]))
  }
}

# create dropdown buttons as list
button_list <- list()
for (i in 1:3) {
  v <- rep(FALSE,3)
  v[i] <- TRUE
  button_list[[i]] <- list(method = "restyle",
                       args = list("visible", as.list(rep(v, each = 4))),
                       label = paste0("COI_mean = ", COI_mean[i]))
}

# finalise layout plto plot
p <- layout(p,
            title = "lambda1 = 10",
            yaxis = list(title = "probability correct assignment", range = c(0,1)),
            xaxis = list(title = "sample size", range = c(0,250)),
            updatemenus = list(
              list(
                yanchor = 'auto',
                buttons = button_list
                )
              )
            )

# print plot
p
```
